---
title: "caEditR: cell-type-resolved RNA-editing deconvolution Example Script"
author: "Haim Krupkin+Claude"
date: "2026-09-13"
output:
  rmarkdown::html_document:
    toc: true
    
vignette: >
  %\VignetteIndexEntry{caEditR: cell-type-resolved RNA-editing deconvolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
print("this is a test of printing capabilites :)")
```

# Introduction
contact hkrupkin@stanford.edu/haim.krupkin@gmail.com if problems
`caEditR`tool stuite helps us deconvolve a bulk RNA-editing-ratio matrix (sites x samples) into
per-cell-type per editing site RNA Editing estimates. Specifically, the suite has in it 3 methods:

1. **`caRD_edit()`** ;reference-based: needs a real sorted per cell type refrence of per cell type per site RNA editing levels.  it incorporates gene expression/site coverage to give better per cell type per RNA editing site estimates.
  
2. **`caNRD_edit()`**; no-reference: self-estimates that reference using math similar to CIBERSOSRTx/TCA directly from the bulk cohort itself using Non-Negative Least Squares(NNLS). it incorporates gene expression/site coverage to give better per cell type per RNA editing site estimates.

3. **`TCA_Like()`**; no refrence: a thin wrapper around the real CRAN `TCA` package. no incoporation of gene expression/site coverage to give better per cell type per RNA editing site estimates.

All three return the same output shape: `$deconvolved`, a named list of per cell type
`sites x samples` matrices.

**Key point : ** All the code is in R is also a wrapper to real code we wrote in python, as inital development was done in python.

# Installation

`caEditR` is installed like any other Bioconductor package:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("caEditR")
```

```{r load}
library(caEditR)
```

```{r ensure-tca, eval = FALSE}
# TCA_Like() wraps the real CRAN 'TCA' package (a Suggests, not a hard
# dependency); if it isn't already installed, install it separately:
caEditR:::.ensure_installed("TCA")
```

# Simulate the cohort

The idea is to generate synthetic data of RNA editing levels. this is primarily useful for benchmarking/methodological development.
But it is also a good function to have in general, and shows us the data format we need to work on real data.

In this following chunk, we create a simplistic example of simulated data.


```{r simulate}
#theta is the "expression", while in methods development phi is expression*proportion in bulk sample=RNA share
#mu is the mean RNA editing levels of a site.

theta_skew <- 5.0       # mean fold-enrichment of the expression of the -dominant cell type per site
theta_skew_sd <- 3.0    # the variability between sites in that fold expression

#this is our actual key command 
cohort <- simulate_reference_and_cohort(
  n_sites = 200, n_celltypes = 6,                       # n_sites matches the validated benchmark's scale
  n_reference_samples = 100, n_deconvolve_samples = 900,
  theta_skew = theta_skew, theta_skew_sd = theta_skew_sd,
  mu_mode = "normal", mu_mean = 0.20, mu_sd = 0.08,      
  mean_coverage = 100.0, concentration = 10.0,           
  seed = 0                                               
)
dim(cohort$bulk_editing)       # 200 sites x 900 samples, to deconvolve
dim(cohort$proportions)        # 900 samples x 6 cell types
names(cohort$ground_truth)     # the TRUE per-cell-type editing value for each of the 900 samples (known only because this is simulated)
head(rownames(cohort$bulk_editing))  # realistic-looking (but entirely synthetic) genomic site ids, chrom:pos:strand
#please take note, that this format is how sites should always be used in the caEditR suite.

#we also decipher the bulk expression at this point.
#first we calculate the actual bulk predicted values from the known theta, then we decipher it using our estimate_theta_nnls
#we need to actually calcualate the true values so our "bulk" measurement is reflective correctly.
#it would be correct to think of theta as coverage/expression.
set.seed(42)
theta_true <- cohort$reference_true$theta
bulk_expr <- (cohort$proportions %*% t(theta_true)) *
  matrix(exp(rnorm(nrow(cohort$proportions) * nrow(theta_true), 0, 0.15)),
         nrow = nrow(cohort$proportions))
bulk_expr <- t(bulk_expr)  # -> sites x samples, matching estimate_theta_nnls()'s convention
dimnames(bulk_expr) <- list(rownames(theta_true), rownames(cohort$proportions))

```
The simulation also creates a simulated reference by sampling a sub population from the same distribution, obviously the larger the sampling (n_reference_samples) the more accurate the refrence captures the population.

```{r reference-quality}
cor(as.vector(cohort$reference_estimated$mu), as.vector(cohort$reference_true$mu))
```

## Estimating theta (relative expression weight)
To estimate the RNA share we use use both the expresion and the proportions, this is similar in concept to TCA/CIBERSORTx. we basically get the per cell type "expression" of each site. if expression values are from bulk .
Theta is "relative expression weight" so it says a relative measure, not exactly directly expression, which is a quantitiy.

```{r estimate-theta}
#we estimate theta
theta_estimated <- estimate_theta_nnls(bulk_expr, cohort$proportions)
#thats an amazingly good theta estimation 0-0.
cor(as.vector(theta_estimated), as.vector(theta_true))  # how good the ESTIMATED theta actually is
```

# Deconvolve the 900 bulk samples, three ways
## caRD_edit() ; using the 100-donor estimated reference

```{r card}
#we create the refrence in the correct format

head(cohort$reference_estimated$mu)
head(cohort$reference_estimated$sigma2)
head(theta_estimated)
reference <- list(
  mu = cohort$reference_estimated$mu,
  sigma2 = cohort$reference_estimated$sigma2,
  theta = theta_estimated   # estimated above via NNLS, NOT the ground truth. so this is needed for our 900 samples.
)
#coverage is importnat for us to know how trust worthy a site is.
out_card <- caRD_edit(cohort$bulk_editing,
                      cohort$coverage,
                      cohort$proportions,
                      reference)
str(out_card)
```

## caNRD_edit() we dont require a refrence 0-0


```{r canrd}
out_canrd <- caNRD_edit(cohort$bulk_editing,
                        cohort$coverage,
                        cohort$proportions,
                         theta_estimated,
                        iterative = TRUE) #iterative true is better, but takes a bit more compute



```
Always check that $diagnostics is ok. if we have enough samples, all should be good.


For more detailed qc:

**n_to_c_ratio;** since we are solving a system of equations , we need the ratio of samples to cell types to solve to be better then 1, and prefererablly better then 3, but generally the more the better.

**marginal_n;**: `TRUE` when `n_to_c_ratio < 3` whether we are below the n needed to solve equations... 
  
**`condition_number`**: how much a small wobble in the observed bulk would influence the rarer cell types. anything below 1000 is ok. if a site is over 1000, we might not be able to trust it for rare cell types.

```{r diagnostics-summary}
head(out_canrd$diagnostics)
summary(out_canrd$diagnostics[, c("n_to_c_ratio", "condition_number")])
```

## TCA_Like() a wrapper for the real CRAN TCA package, proportion-only mixing


```{r tca}
#TCA requires input data to have variance, so ew check it for variance, we know it has variance since we simulate
#but generally its good to double check things, and when this is ran on real data, u need to do this step.

bulk_nz <- cohort$bulk_editing[apply(cohort$bulk_editing, 1, var) > 1e-8, , drop = FALSE]
out_tca <- TCA_Like(bulk_nz, cohort$proportions)
```

# Score all three against the known ground truth

Because this is "idealized" simulated data, we know the ground actual true per-cell-type per editing site value for every one of the 900 deconvolved simulated samples. On this simulated data, lets compare all three methods' accuracy:

**Note on `NA`s:** `caRD_edit()`/`caNRD_edit()` deliberately return `NA` for a (site, cell type) pair when that cell type's expression signal at that site is too close to caEditR's own detection floor to identify at all (see `?caRD_edit`) -- reporting nothing is safer than a numerically unstable guess. `cor(..., use = "complete.obs")` below scores only the cell/sample pairs that were actually estimated by both methods being compared.

```{r score}
celltypes <- names(cohort$ground_truth)
r2 <- data.frame(
  celltype = celltypes,
  caRD_edit = sapply(celltypes, function(ct) cor(as.vector(out_card$deconvolved[[ct]]), as.vector(cohort$ground_truth[[ct]]), use = "complete.obs")^2),
  caNRD_edit = sapply(celltypes, function(ct) cor(as.vector(out_canrd$deconvolved[[ct]]), as.vector(cohort$ground_truth[[ct]]), use = "complete.obs")^2),
  TCA_Like = sapply(celltypes, function(ct) {
    gt <- cohort$ground_truth[[ct]][rownames(bulk_nz), ]
    cor(as.vector(out_tca$deconvolved[[ct]]), as.vector(gt))^2
  })
)
print(r2, row.names = FALSE)
```
Lets plot the acutall accuracy
```{r plot-r2, fig.alt = "Bar plot of Pearson R-squared (deconvolved estimate vs. simulated ground truth) for each of 6 cell types, grouped by method (caRD_edit, caNRD_edit, TCA_Like)."}
r2_long <- reshape(r2, direction = "long", varying = list(2:4),
                    v.names = "r2", timevar = "method", times = colnames(r2)[2:4])
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(r2_long, ggplot2::aes(x = celltype, y = r2, fill = method)) +
    ggplot2::geom_col(position = "dodge") +
    ggplot2::labs(y = expression(Pearson~R^2~"(estimate vs truth)",), x = NULL,
                  fill="Method") +
    ggplot2::theme_bw()+
    ggplot2::theme(legend.position="bottom")
} else {
  barplot(t(as.matrix(r2[, -1])), beside = TRUE, names.arg = r2$celltype,
          legend.text = colnames(r2)[-1], ylab = "Pearson R^2 (estimate vs. true)")
}
```

## Scatter plots: true vs. deconvolved, all three methods

The R^2 bars above summarize accuracy per cell type; the scatter plots show us per site what is happening. If you look, you will see TCA's inherent misstake. which is, if a cell type that has high expression, it will overestimate its RNA editing levels, because it will say "oh we have all this excess RNA editing" because it doesn't account for a specific cell types expression, thus over estimating its RNA editing. A similar thing could happen the opposite way, where it will underestimates a cell type with low expression.

(`caRD_edit()`/`caNRD_edit()`'s panels below will show noticeably fewer points than `TCA_Like`'s -- and a "Removed N rows containing missing values" warning -- because of the same `NA`-for-unidentifiable-signal behavior noted above; `TCA_Like` has no equivalent identifiability gating, so it always plots a point for every cell/sample.)

```{r scatter-data}
make_long <- function(out, gt, celltypes, sample_subset = NULL) {
  rows <- lapply(celltypes, function(ct) {
    true_vals <- gt[[ct]]
    est_vals <- out[[ct]]
    if (!is.null(sample_subset)) {
      true_vals <- true_vals[, sample_subset, drop = FALSE]
    }
    common_samples <- intersect(colnames(true_vals), colnames(est_vals))
    data.frame(celltype = ct, true = as.vector(true_vals[, common_samples]),
               estimate = as.vector(est_vals[, common_samples]))
  })
  do.call(rbind, rows)
}

scatter_df <- rbind(
  cbind(method = "caRD_edit", make_long(out_card$deconvolved, cohort$ground_truth, celltypes)),
  cbind(method = "caNRD_edit", make_long(out_canrd$deconvolved, cohort$ground_truth, celltypes)),
  cbind(method = "TCA_Like", make_long(out_tca$deconvolved, cohort$ground_truth, celltypes))
)
```

```{r plot-scatter, fig.width = 9, fig.height = 3.2}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(scatter_df, ggplot2::aes(x = true, y = estimate, color = celltype)) +
    ggplot2::geom_point(alpha = 0.15, size = 0.6) +
    ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
    ggplot2::facet_wrap(~method, nrow = 1) +
    ggplot2::coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    ggplot2::labs(x = "True (simulated) editing ratio", y = "Deconvolved estimate") +
    ggplot2::theme_minimal()
} else {
  par(mfrow = c(1, 3))
  for (m in unique(scatter_df$method)) {
    d <- scatter_df[scatter_df$method == m, ]
    plot(d$true, d$estimate, pch = 16, cex = 0.3, col = adjustcolor("steelblue", alpha.f = 0.2),
         xlim = c(0, 1), ylim = c(0, 1), main = m, xlab = "True", ylab = "Estimate")
    abline(0, 1, lty = 2)
  }
}
print("done creating the plots :) ")
```



# Real data:
**GSE64655 Ottoboni et al** has 8 real bulk PBMC samples

```{r gse64655-load}
print("loading real data")
extdata <- system.file("extdata", package = "caEditR")
read_matrix <- function(f) as.matrix(read.csv(file.path(extdata, f), row.names = 1, check.names = FALSE))

real_bulk <- read_matrix("gse64655_bulk_editing_ratios.csv")       # 107 real sites x 8 real bulk samples
real_coverage <- read_matrix("gse64655_bulk_coverage.csv")         # REAL per-site coverage this time, not a placeholder
real_gene_counts <- read_matrix("gse64655_bulk_gene_counts.csv")   # REAL featureCounts, same 8 real bulk samples
real_proportions <- read_matrix("gse64655_proportions.csv")        # real, MuSiC-estimated (pooled, see disclosure above)
real_reference <- list(
  mu = read_matrix("gse64655_reference_mu.csv"),
  sigma2 = read_matrix("gse64655_reference_sigma2.csv"),
  theta = read_matrix("gse64655_reference_theta.csv")
)
dim(real_bulk)
print("done loading real data")
```

## Getting proportions the honest way, on this real data
If needed, we also enable cell type proportion estimation in this package. 
Same idea as papers before, given real per-sample gene expression and a real,
independent signature matrix (Monaco et al. 2019, `blood_signature_matrix.csv`, estimate proportions via NNLS.

```{r signature-matrix-proportions}
print("loading expression signature matrix")
sig <- as.matrix(read.csv(file.path(extdata, "blood_signature_matrix.csv"), row.names = 1))
sig_proportions <- estimate_proportions_signature_matrix(real_gene_counts, sig)
sig_proportions
print("Done loading expression signature matrix")

```

## The standardized site id format



```{r format-site-id}
format_site_id("10", 100232436, "-")        # -> "10:100232436:-"
format_site_id("chr10", 100232436, "-")     # "chr" prefix is stripped automatically
format_site_id("1", 12831014)               # strand unspecified -> no trailing ":*"
```

## Mapping RNA editing sites to the genes they are found in, and deriving coverage from expression, as a proxy.

Exactly the same functions as before, just pointed at GSE64655 - genome` build parameter is required **GRCh38/hg38**.
By default map sites to genes uses Ensembl release 110

This section (and "Passing `expression` directly instead of a separate
coverage step" below) needs the real Bioconductor UCSC genome-annotation
packages (`GenomicFeatures`, `TxDb.Hsapiens.UCSC.hg38.knownGene`,
`org.Hs.eg.db`, `AnnotationDbi`). These are listed in caEditR's own
`Suggests`, not `Imports` -- a plain `BiocManager::install("caEditR")`
does NOT install them, so this vignette checks for them explicitly and
skips gracefully (rather than hard-failing) if they are not present. If
this chunk is skipped for you, install them with:
`BiocManager::install(c("GenomicFeatures", "TxDb.Hsapiens.UCSC.hg38.knownGene", "org.Hs.eg.db", "AnnotationDbi"))`.

```{r check-genome-annotation}
has_genome_annotation <- requireNamespace("GenomicFeatures", quietly = TRUE) &&
  requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
if (!has_genome_annotation) {
  message("Skipping gene-mapping/expression-derived-coverage sections: ",
          "one or more of GenomicFeatures/TxDb.Hsapiens.UCSC.hg38.knownGene/",
          "org.Hs.eg.db/AnnotationDbi is not installed. The rest of the ",
          "vignette does not depend on this and will still run using the ",
          "real, directly-measured coverage already loaded above.")
}
```

```{r gse64655-map-genes}
# The if() here (not just this chunk's `eval=` option) is deliberate: it
# makes the skip work identically whether this .Rmd is knitted OR run
# interactively line-by-line/chunk-by-chunk in RStudio -- `eval=` chunk
# options are a knitr-render-time-only mechanism and are silently ignored
# when you just run the code directly, which is exactly how this was
# caught not actually protecting anyone running it that way.
if (has_genome_annotation) {
  site_to_gene <- map_sites_to_genes(rownames(real_bulk), genome = "hg38")
  cat(sprintf("%d/%d real sites mapped to a real gene on hg38\n",
              sum(!is.na(site_to_gene$ensembl_gene_id)), nrow(site_to_gene)))

  coverage_from_expr <- build_coverage_from_expression(rownames(real_bulk), real_gene_counts, genome = "hg38")
  head(coverage_from_expr, 3)
}
```

## Deconvolving real data, in the 3 ways

```{r gse64655-deconvolve}
out_card <- caRD_edit(real_bulk, real_coverage, real_proportions, real_reference)

# iterative=TRUE, ridge_frac=0.1 -- the exact configuration this project's own
# validated real-data figure found necessary; real N=8 data is badly
# ill-conditioned (median condition number ~1.7 MILLION unregularized).
out_canrd <- caNRD_edit(real_bulk, real_coverage, real_proportions, real_reference$theta,
                         iterative = TRUE, ridge_frac = 0.1)

bulk_nz <- real_bulk[apply(real_bulk, 1, var) > 1e-8, , drop = FALSE]
out_tca <- TCA_Like(bulk_nz, real_proportions)

# The honest part: with only 8 real samples against 6 cell types (need >=18
# to be "reliable"), caNRD_edit's OWN diagnostics correctly flag this cohort
# as marginal for essentially every site that gets an estimate at all
# (na.rm=TRUE because a couple of sites are excluded outright now --
# floor-clamped/negligible-signal cell types -- rather than flagged
# marginal; see `status`/`excluded_celltypes` for those):
mean(out_canrd$diagnostics$marginal_n, na.rm = TRUE)
sum(out_canrd$diagnostics$status == "insufficient_identifiable_celltypes")
```




```{r gse64655-score}
celltypes <- colnames(real_proportions)  # GSE64655's own 6 cell types, NOT the earlier simulated section's
real_r2 <- do.call(rbind, lapply(celltypes, function(ct) {
  gt <- read_matrix(paste0("gse64655_ground_truth_", ct, ".csv"))
  genuine <- as.matrix(read.csv(file.path(extdata, paste0("gse64655_genuine_", ct, ".csv")),
                                 row.names = 1, check.names = FALSE)) == "True"
  genuine[is.na(genuine)] <- FALSE
  n <- sum(genuine)
  r2 <- function(out) {
    est <- out$deconvolved[[ct]][rownames(genuine), colnames(genuine)]
    if (n >= 3) cor(est[genuine], gt[rownames(genuine), colnames(genuine)][genuine], use = "complete.obs")^2 else NA_real_
  }
  data.frame(celltype = ct, n_genuine = n,
             caRD_edit = r2(out_card), caNRD_edit = r2(out_canrd), TCA_Like = r2(out_tca))
}))
print(real_r2, row.names = FALSE)
```

## How many (site, sample) pairs did we actually get an estimate for?

The R^2 table above only scores what got *estimated*; it says nothing about
*how much* of the real data that covers. Cell types differ hugely here --
worth checking directly rather than assuming "we ran it on 107 sites" means
every cell type got 107 sites' worth of answers.

```{r gse64655-coverage}
n_total <- nrow(real_bulk) * ncol(real_bulk)  # sites x samples
coverage_summary <- do.call(rbind, lapply(colnames(real_proportions), function(ct) {
  data.frame(
    celltype = ct,
    pct_estimated_caRD = 100 * sum(!is.na(out_card$deconvolved[[ct]])) / n_total,
    pct_estimated_caNRD = 100 * sum(!is.na(out_canrd$deconvolved[[ct]])) / n_total,
    mean_real_proportion_pct = 100 * mean(real_proportions[, ct])
  )
}))
print(coverage_summary, row.names = FALSE, digits = 2)

cat("\nSite-level status (out of", nrow(real_bulk), "real sites):\n")
print(table(out_card$diagnostics$status))
```

**Neutrophils get 0% coverage and only 2 "genuine" ground-truth pairs --
confirmed directly, not a bug**: GSE64655 is a *PBMC* (peripheral blood
**mononuclear** cell) dataset. PBMC isolation (density-gradient
centrifugation) specifically removes granulocytes, including neutrophils,
by design -- that is the entire point of a PBMC prep. Real neutrophil
proportion across all 8 samples here averages `r round(100 * mean(real_proportions[, "Neutrophils"]), 2)`%.
With essentially no real neutrophil signal in these particular samples,
there is nothing for either method to identify, and nothing reliable to
build "genuine" validation ground truth from -- this is a fact about the
input data, not about `caRD_edit()`/`caNRD_edit()`'s gating. B cells and
DCs are also low-coverage here for the more familiar reason: they are
genuinely minor populations in blood, so most sites' host genes simply
are not expressed distinctly enough in them to be identifiable from bulk
(see `?caRD_edit`'s `min_mean_phi` for the exact mechanism).

```{r gse64655-plot-coverage, fig.alt = "Scatter plot of the percent of (site, sample) pairs caRD_edit() actually estimated for each of the 6 real GSE64655 cell types (y-axis) against that cell type's real mean proportion in the cohort (x-axis), one labeled point per cell type, showing coverage tracks real abundance."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(coverage_summary, ggplot2::aes(x = mean_real_proportion_pct, y = pct_estimated_caRD, label = celltype)) +
    ggplot2::geom_point(size = 2.5, color = "steelblue") +
    ggplot2::geom_text(vjust = -0.8, size = 3.5) +
    ggplot2::labs(x = "Real mean cell-type proportion (%)", y = "% of (site, sample) pairs estimated (caRD_edit)") +
    ggplot2::coord_cartesian(xlim = c(0, max(coverage_summary$mean_real_proportion_pct) * 1.15),
                              ylim = c(0, 100)) +
    ggplot2::theme_bw()
} else {
  plot(coverage_summary$mean_real_proportion_pct, coverage_summary$pct_estimated_caRD,
       xlab = "Real mean cell-type proportion (%)", ylab = "% of (site, sample) pairs estimated (caRD_edit)",
       pch = 16, col = "steelblue", ylim = c(0, 100))
  text(coverage_summary$mean_real_proportion_pct, coverage_summary$pct_estimated_caRD,
       labels = coverage_summary$celltype, pos = 3)
}
```

```{r gse64655-plot-r2, fig.alt = "Bar plot of Pearson R-squared (deconvolved estimate vs. real, directly-measured ground truth) for each of the 6 real GSE64655 cell types, grouped by method (caRD_edit, caNRD_edit, TCA_Like)."}
real_r2_long <- reshape(real_r2, direction = "long", varying = list(3:5),
                         v.names = "r2", timevar = "method", times = colnames(real_r2)[3:5])
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(real_r2_long, ggplot2::aes(x = celltype, y = r2, fill = method)) +
    ggplot2::geom_col(position = "dodge") +
    ggplot2::labs(y = expression(Pearson~R^2~"(estimate vs. real ground truth)"), x = NULL,
                  fill = "Method") +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "bottom")
} else {
  barplot(t(as.matrix(real_r2[, 3:5])), beside = TRUE, names.arg = real_r2$celltype,
          legend.text = colnames(real_r2)[3:5], ylab = "Pearson R^2 (estimate vs. real ground truth)")
}
```

**Interpretation, stated honestly**:

## Passing `expression` directly instead of a separate coverage step

This also needs the same optional genome-annotation packages checked
above (`expression`-derived coverage internally calls
`build_coverage_from_expression()`, which needs them to map sites to
genes) -- skipped gracefully here too if they are not installed.

```{r gse64655-integrated}
# Same reasoning as the previous chunk: a real if(), not just eval=, so
# this is actually skipped when run interactively too.
if (has_genome_annotation) {
  #if wanted, u can also do this format instead:
  out_card_from_expr <- caRD_edit(
    real_bulk, proportions = real_proportions, reference = real_reference,
    expression = real_gene_counts, genome = "hg38"
  )
  head(out_card_from_expr$deconvolved$Neutrophils, 3)
}
```

`coverage` (a real matrix) and `expression` are mutually exclusive ways to supply the same argument. if you have real per-site coverage, always prefer it (`coverage` silently wins if both are given, with a `message()` saying so); `expression`derived coverage is a proxy for when you don't have coverage.

# Session info
s
```{r sessionInfo}
sessionInfo()
```
