---
title: "caEditR: complete workflow on simulated data (deconvolution, cell-type edQTLs, genotype-informed reconstruction)"
author: "Haim Krupkin+Claude"
date: "2026-09-30"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{caEditR: complete workflow on simulated data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 4.5)
```

# What this vignette covers

Every exported caEditR function, on simulated data with known truth, in the order of a real analysis:

| step | functions |
|---|---|
| site ids and per-site annotation | `format_site_id()`, `map_sites_to_genes()`, `build_coverage_from_expression()` |
| simulating cohorts | `simulate_reference()`, `simulate_proportions()`, `simulate_true_editing()`, `simulate_bulk()`, `simulate_reference_and_cohort()`, `simulate_bulk_expression()`, `simulate_edqtl_cohort()` |
| inputs | `estimate_theta_nnls()`, `estimate_proportions_signature_matrix()`, `load_reference()`, `compute_effective_weights()`, `binomial_tau2()` |
| deconvolution (per-sample matrices) | `caRD_edit()`, `caNRD_edit()` (and earlier versions `caNRDv0.5_edit()`, `caNRDv0_edit()`), `TCA_Like()`, `estimate_no_reference_params()`, `deconvolve_site()` |
| cell-type edQTLs | `celltype_edqtl()`, `caNRD_editQTL()` (engines `"fast"`, `"reference"`, `"scan"`), `caNRD_editQTL_bootstrap()` |
| genotype-informed matrices | `caNRD_editQTL_shrink()`, `caNRD_joint_reconstruction()` |

It runs with only `module load R/4.3.3` (plus the caEditR package itself): `caRD_edit()` / `caNRD_edit()` call the
package's bundled Python core and find a Python with numpy/scipy on their own. The two genome-annotation functions
need the Bioconductor annotation packages (`GenomicFeatures` and a UCSC `TxDb`); their chunks run only when those are
installed.

```{r load}
library(caEditR)
Sys.setenv(PYTHONNOUSERSITE = "1")   # the child Python used by caRD_edit()/caNRD_edit() ignores ~/.local packages
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

# 1. Site ids and per-site annotation

Every caEditR function identifies an RNA-editing site by one standardized id, `"chrom:pos:strand"` (no `"chr"`
prefix, 1-based position, optional strand). Build them with `format_site_id()`:

```{r site-ids}
format_site_id("10", 100232436, "-")
format_site_id("chr1", c(12831014, 12831020), c("+", "+"))
format_site_id("X", 73820651)           # strand unknown -> "X:73820651"
```

To use host-gene **expression** as a proxy for read **coverage** (when coverage is not available), map sites to
their host genes and turn a gene x sample expression matrix into a site x sample coverage matrix.
`caRD_edit()` / `caNRD_edit()` can also do this themselves through their `expression =` argument.

```{r annotation}
have_annot <- requireNamespace("GenomicFeatures", quietly = TRUE) &&
  requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE)
if (have_annot) {
  sites <- format_site_id(c("10", "10"), c(100232436, 100232925), c("-", "-"))
  print(map_sites_to_genes(sites, genome = "hg38"))
  expr <- matrix(c(120, 80, 45, 60), 2, 2, dimnames = list(c("ENSG00000107954", "ENSG00000000003"), c("s1", "s2")))
  print(build_coverage_from_expression(sites, expr, genome = "hg38"))
} else {
  message("GenomicFeatures / TxDb.Hsapiens.UCSC.hg38.knownGene not installed: map_sites_to_genes() and ",
          "build_coverage_from_expression() are skipped here (they need the Bioconductor annotation packages).")
}
```

# 2. Deconvolution: per-sample cell-type editing matrices

## 2.1 Simulating a cohort from its building blocks

`simulate_reference()` draws per-site, per-cell-type editing means `mu`, variances `sigma2` and host-gene expression
weights `theta`; `simulate_proportions()` draws cell-type proportions; `simulate_true_editing()` draws each sample's
true cell-type editing; `simulate_bulk()` mixes them by RNA share and draws reads.

```{r building-blocks}
ref0 <- simulate_reference(n_sites = 4, n_celltypes = 3, seed = 1)
p0 <- simulate_proportions(n_samples = 6, composition = c(0.6, 0.3, 0.1), seed = 1)
e0 <- simulate_true_editing(ref0$mu, ref0$sigma2, n_samples = 6, seed = 1)   # samples x sites x cell types
b0 <- simulate_bulk(e0, p0, ref0$theta, mean_coverage = 50, seed = 1)
round(b0$e_bulk_obs, 3)                                                        # sites x samples observed bulk editing
```

`simulate_reference_and_cohort()` does all of this at once, with named matrices, a reference set of donors and a
cohort to deconvolve:

```{r simulate-cohort}
cohort <- simulate_reference_and_cohort(n_sites = 150, n_celltypes = 6, n_reference_samples = 100,
                                        n_deconvolve_samples = 600, theta_skew = 5, theta_skew_sd = 3,
                                        mu_mode = "normal", mu_mean = 0.20, mu_sd = 0.08, mean_coverage = 100,
                                        concentration = 10, seed = 0)
dim(cohort$bulk_editing); dim(cohort$proportions); head(rownames(cohort$bulk_editing), 3)
```

## 2.2 Inputs: expression weights, proportions, references

**theta** (relative host-gene expression per cell type) is estimated from bulk expression and proportions by
non-negative least squares. `simulate_bulk_expression()` gives the bulk expression of each site's host gene:

```{r theta}
bulk_expr <- simulate_bulk_expression(cohort$proportions, cohort$reference_true$theta, noise_sdlog = 0.15, seed = 42)
theta_hat <- estimate_theta_nnls(bulk_expr, cohort$proportions)
cor(as.vector(theta_hat), as.vector(cohort$reference_true$theta))
```

**Proportions** can be estimated from bulk gene counts and a cell-type signature matrix:

```{r proportions}
set.seed(2)
sig <- matrix(stats::rgamma(60 * 6, 2), 60, 6, dimnames = list(paste0("gene", 1:60), colnames(cohort$proportions)))
counts <- sig %*% t(cohort$proportions[1:50, ]) * 1000
p_hat <- estimate_proportions_signature_matrix(counts, sig)
cor(as.vector(as.matrix(p_hat)), as.vector(cohort$proportions[1:50, ]))
```

**RNA shares** (`phi`, proportions x expression, renormalized) and the binomial read-sampling variance `tau2`:

```{r phi-tau2}
phi1 <- compute_effective_weights(cohort$proportions, theta_hat[1, ], mode = "phi")   # site 1, every sample (rows)
round(head(phi1, 3), 3)
round(head(binomial_tau2(cohort$bulk_editing[1, ], cohort$coverage[1, ])), 6)
```

`load_reference()` loads the reference bundled with the package (real sorted-cell editing; 1,495 sites x 6 cell
types):

```{r bundled-reference}
bundled <- load_reference(); sapply(bundled, dim)
```

## 2.3 Three deconvolution methods

```{r deconvolve}
reference <- list(mu = cohort$reference_estimated$mu, sigma2 = cohort$reference_estimated$sigma2, theta = theta_hat)
out_card  <- caRD_edit(cohort$bulk_editing, cohort$coverage, cohort$proportions, reference, theta_floor = 1e-3)
out_canrd <- caNRD_edit(cohort$bulk_editing, cohort$coverage, cohort$proportions, theta_hat, theta_floor = 1e-3, iterative = TRUE)
bulk_nz   <- cohort$bulk_editing[apply(cohort$bulk_editing, 1, stats::var) > 1e-8, , drop = FALSE]
out_tca   <- TCA_Like(bulk_nz, cohort$proportions)          # CRAN TCA's per-sample estimates (proportions only)
summary(out_canrd$diagnostics[, c("n_to_c_ratio", "condition_number")])
```

Scored against the known truth (R squared per cell type; `NA` estimates are cell types caEditR declines to estimate):

```{r score}
cts <- names(cohort$ground_truth)
r2 <- data.frame(celltype = cts,
  caRD_edit  = sapply(cts, function(h) cor(as.vector(out_card$deconvolved[[h]]), as.vector(cohort$ground_truth[[h]]), use = "complete.obs")^2),
  caNRD_edit = sapply(cts, function(h) cor(as.vector(out_canrd$deconvolved[[h]]), as.vector(cohort$ground_truth[[h]]), use = "complete.obs")^2),
  TCA_Like   = sapply(cts, function(h) cor(as.vector(out_tca$deconvolved[[h]]), as.vector(cohort$ground_truth[[h]][rownames(bulk_nz), ]))^2))
print(r2, row.names = FALSE, digits = 3)
```

The building blocks of `caNRD_edit()` for one site: estimate the no-reference parameters, then the per-sample
posterior means.

```{r low-level}
s <- 1
phi_s <- compute_effective_weights(cohort$proportions, theta_hat[s, ], mode = "phi")
tau2_s <- binomial_tau2(cohort$bulk_editing[s, ], cohort$coverage[s, ])
par_s <- estimate_no_reference_params(cohort$bulk_editing[s, ], phi_s, tau2_s, iterative = TRUE)
# deconvolve_site() is the per-sample posterior mean (one sample at a time); first 5 samples:
z_s <- t(vapply(1:5, function(i) deconvolve_site(cohort$bulk_editing[s, i], phi_s[i, ], tau2_s[i], par_s$mu_hat, par_s$sigma2_hat),
                numeric(ncol(phi_s))))
round(par_s$mu_hat, 3); round(z_s, 3)
```

`caNRD_edit()` estimates each site's parameters by maximum likelihood with the read-sampling variance taken from the
model-predicted bulk level ("variance-fixed" caNRD). The previous estimator, which used each donor's observed ratio
and is biased downward at low-read donors, is kept as `caNRDv0.5_edit()` (alias `caNRDv2_edit()`); `caNRDv0_edit()`
is the original ungated version and `caRDv0_edit()` the original `caRD_edit()`. Comparing the two caNRD estimators
against the truth on this cohort:

```{r canrd-versions}
out_v05 <- caNRDv0.5_edit(cohort$bulk_editing, cohort$coverage, cohort$proportions, theta_hat, theta_floor = 1e-3, iterative = TRUE)
data.frame(celltype = cts,
  caNRD_edit = sapply(cts, function(h) cor(as.vector(out_canrd$deconvolved[[h]]), as.vector(cohort$ground_truth[[h]]), use = "complete.obs")^2),
  caNRDv0.5_edit = sapply(cts, function(h) cor(as.vector(out_v05$deconvolved[[h]]), as.vector(cohort$ground_truth[[h]]), use = "complete.obs")^2),
  bias_caNRD_edit_pp = sapply(cts, function(h) 100 * mean(out_canrd$deconvolved[[h]] - cohort$ground_truth[[h]], na.rm = TRUE)),
  bias_caNRDv0.5_pp = sapply(cts, function(h) 100 * mean(out_v05$deconvolved[[h]] - cohort$ground_truth[[h]], na.rm = TRUE)))
```

# 3. Cell-type edQTLs

## 3.1 An edQTL cohort with known genetic effects

`simulate_edqtl_cohort()` simulates donors, genotypes (several cis variants per site: one causal, two in LD with
it, the rest independent), realistic read depth, host genes not expressed in some cell types, and sites with an effect
in one cell type, a shared effect, or none:

```{r edqtl-cohort}
eq <- simulate_edqtl_cohort(n_donors = 1500, n_sites_per_type = 20, n_variants = 5, beta = 0.06, mean_coverage = 60, seed = 7)
table(eq$truth$site_type)
dim(eq$genotypes); head(eq$pairs, 3)
summary(as.vector(eq$coverage))
theta_eq <- estimate_theta_nnls(eq$bulk_expression, eq$proportions)    # expression weights from bulk expression
```

## 3.2 Testing: celltype_edqtl() and caNRD_editQTL()

Both fit the genotype x RNA-share mixture model jointly across cell types on bulk editing, so a genetic effect is
attributed to the cell type whose RNA share it follows (per-sample matrices tested one cell type at a time leak the
signal into the others). `caNRD_editQTL()` uses caNRD's latent-variable variance model and likelihood-ratio tests;
its engines compute the same estimator (`"fast"`, `"reference"`) or, for genome-wide cis scans, test all variants of
a site at that site's genotype-free variance and re-fit the lead variant exactly (`"scan"`).

```{r edqtl-test}
causal <- data.frame(site_id = names(eq$truth$causal_variant), variant_id = unname(eq$truth$causal_variant))
t_ct   <- system.time(r_ct   <- celltype_edqtl(eq$bulk_editing, eq$genotypes, eq$proportions, theta_eq, theta_floor = 1e-3,
                                               pairs = causal, coverage = eq$coverage))[["elapsed"]]
t_fast <- system.time(r_fast <- caNRD_editQTL(eq$bulk_editing, eq$genotypes, eq$proportions, theta_eq, theta_floor = 1e-3,
                                              pairs = causal, coverage = eq$coverage))[["elapsed"]]
t_scan <- system.time(r_scan <- caNRD_editQTL(eq$bulk_editing, eq$genotypes, eq$proportions, theta_eq, theta_floor = 1e-3,
                                              pairs = eq$pairs, coverage = eq$coverage, engine = "scan"))[["elapsed"]]
r_ref  <- caNRD_editQTL(eq$bulk_editing[1:2, ], eq$genotypes, eq$proportions, theta_eq[1:2, ], theta_floor = 1e-3,
                        pairs = causal[1:2, ], coverage = eq$coverage[1:2, ], engine = "reference")
c(celltype_edqtl_s = t_ct, fast_s = t_fast, scan_all_pairs_s = t_scan)
all.equal(r_ref$beta, r_fast$beta[r_fast$site_id %in% causal$site_id[1:2]], tolerance = 1e-3)
```

Calls at the causal variants (p < 0.001) against the truth: the true cell type should be found, and the others
rarely called.

```{r edqtl-calls}
truth_long <- data.frame(site_id = rep(rownames(eq$truth$beta), ncol(eq$truth$beta)),
                         celltype = rep(colnames(eq$truth$beta), each = nrow(eq$truth$beta)),
                         beta_true = as.vector(eq$truth$beta))
score <- function(r, name) {
  m <- merge(r[r$status == "tested", c("site_id", "celltype", "beta", "p")], truth_long, by = c("site_id", "celltype"))
  data.frame(method = name, power_true_celltype = mean(m$p[m$beta_true != 0] < 1e-3),
             wrong_celltype_rate = mean(m$p[m$beta_true == 0] < 1e-3), cor_beta = cor(m$beta, m$beta_true))
}
rbind(score(r_ct, "celltype_edqtl"), score(r_fast, "caNRD_editQTL fast"),
      score(r_scan[r_scan$variant_id %in% causal$variant_id, ], "caNRD_editQTL scan (causal variants)"))
```

The scan engine tests every variant; the lead variant of each site (re-fitted exactly) is usually the causal one or a
variant in LD with it:

```{r scan-leads}
leads <- unique(r_scan[r_scan$refined, c("site_id", "variant_id", "p_site")])
leads$role <- eq$truth$variants$role[match(leads$variant_id, eq$truth$variants$variant_id)]
leads$site_type <- eq$truth$site_type[leads$site_id]
table(site_type = leads$site_type, lead_variant = leads$role)
```

Bootstrap fits (resampling donors) give parameter uncertainty for reconstruction intervals:

```{r bootstrap}
boot <- caNRD_editQTL_bootstrap(eq$bulk_editing, eq$genotypes, eq$proportions, theta_eq, theta_floor = 1e-3,
                                pairs = causal, coverage = eq$coverage, n_boot = 4, seed = 1)
length(boot); sd(sapply(boot, function(f) f$beta[1]))
```

# 4. Genotype-informed cell-type editing matrices

`caNRD_joint_reconstruction()` reconstructs per-donor cell-type editing like `caNRD_edit()`, but uses the joint
genotype effects of a `caNRD_editQTL()` fit (one variant per site) so the genetic signal is placed in the cell type
it belongs to before the remaining bulk signal is shared out. `caNRD_editQTL_shrink()` first shrinks the betas
(multivariate point-normal empirical Bayes, the recommended setting), which reduces the false genetic structure that
noisy betas would otherwise write into unaffected cell types. The discovery results (`r_fast`) are not modified.

```{r reconstruct}
shrunk <- caNRD_editQTL_shrink(r_fast, eq$bulk_editing, eq$genotypes, eq$proportions, theta_eq, theta_floor = 1e-3,
                               coverage = eq$coverage)
attr(shrunk, "prior")
rec <- caNRD_joint_reconstruction(eq$bulk_editing, eq$genotypes, eq$proportions, theta_eq, theta_floor = 1e-3,
                                  fit = shrunk, coverage = eq$coverage, boot_fits = boot)
table(rec$diagnostics$status)
plain <- caNRD_edit(eq$bulk_editing, eq$coverage, eq$proportions, theta_eq, theta_floor = 1e-3, iterative = TRUE)
```

The shrinkage prior is learned from all sites together: per cell type, the fraction of sites with no effect (`pi0`)
and the typical size of real effects (`t2`). It needs many sites to be reliable (here 73-97 per cell type; genome-wide
data has thousands); with only a handful of sites it can mis-shrink real effects.

One Monocyte-only site, chosen by a stated rule (as one would pick an example in real data): among Monocyte-only sites
whose host gene is expressed in every cell type, the one with the strongest Monocyte evidence in the joint fit
(smallest `p`). Genetic slope per cell type in the truth, the plain caNRD matrix and the genotype-informed matrix
(percentage points of editing per allele):

```{r one-site}
mono <- names(eq$truth$site_type)[eq$truth$site_type == "only_Monocytes" & rowSums(eq$theta > 0) == ncol(eq$theta)]
fm <- r_fast[r_fast$site_id %in% mono & r_fast$celltype == "Monocytes", ]
site <- fm$site_id[which.min(fm$p)]
c(site = site, monocyte_p = signif(min(fm$p, na.rm = TRUE), 2))
g <- eq$genotypes[eq$truth$causal_variant[site], ]
slope <- function(z) if (all(is.na(z))) NA_real_ else 100 * stats::coef(stats::lm(z ~ g))[2]
data.frame(celltype = colnames(eq$proportions),
           expressed = eq$theta[site, ] > 0,
           truth = sapply(colnames(eq$proportions), function(h) slope(eq$truth$latent[[h]][site, ])),
           caNRD_edit = sapply(colnames(eq$proportions), function(h) slope(plain$deconvolved[[h]][site, ])),
           joint_reconstruction = sapply(colnames(eq$proportions), function(h) slope(rec$reconstructed[[h]][site, ])))
```

Across all single-cell-type sites: how much of the true effect each matrix keeps in the right cell type, and how
much genetic slope it puts into cell types without an effect (absolute, so errors in opposite directions don't
cancel). These slopes are diagnostics of the matrices, not association tests: the genotype-informed matrix contains
the fitted genotype effect by construction, so edQTLs are tested with `caNRD_editQTL()`, never on these matrices.

```{r leakage}
single <- names(eq$truth$site_type)[grepl("^only_", eq$truth$site_type)]
lk <- do.call(rbind, lapply(single, function(s) {
  gg <- eq$genotypes[eq$truth$causal_variant[s], ]; tgt <- sub("^only_", "", eq$truth$site_type[s])
  do.call(rbind, lapply(colnames(eq$proportions)[eq$theta[s, ] > 0], function(h) {
    sl <- function(z) if (all(is.na(z))) NA_real_ else stats::coef(stats::lm(z ~ gg))[2]
    data.frame(site = s, celltype = h, affected = h == tgt, true_beta = eq$truth$beta[s, h],
               caNRD_edit = sl(plain$deconvolved[[h]][s, ]), joint = sl(rec$reconstructed[[h]][s, ]))
  }))
}))
with(lk, data.frame(
  matrix = c("caNRD_edit", "joint_reconstruction"),
  effect_kept = c(mean(caNRD_edit[affected], na.rm = TRUE), mean(joint[affected], na.rm = TRUE)) / 0.06,
  fake_effect_pp_per_allele = 100 * c(mean(abs(caNRD_edit[!affected]), na.rm = TRUE), mean(abs(joint[!affected]), na.rm = TRUE))))
```

Interval widths from the bootstrap fits (`total_sd`) vs. the parameter-conditional ones (`conditional_sd`):

```{r intervals}
h <- "Monocytes"
c(conditional_sd = median(rec$conditional_sd[[h]], na.rm = TRUE), total_sd = median(rec$total_sd[[h]], na.rm = TRUE))
```

```{r plot-site, fig.alt = "Editing by genotype for one Monocyte-only site: truth, caNRD_edit and genotype-informed reconstruction."}
if (has_ggplot) {
  d <- do.call(rbind, lapply(c("truth", "caNRD_edit", "joint_reconstruction"), function(src) {
    M <- switch(src, truth = eq$truth$latent, caNRD_edit = plain$deconvolved, joint_reconstruction = rec$reconstructed)
    do.call(rbind, lapply(colnames(eq$proportions)[eq$theta[site, ] > 0], function(hh)
      data.frame(source = src, celltype = hh, genotype = factor(g), editing = 100 * M[[hh]][site, ])))
  }))
  d$source <- factor(d$source, levels = c("truth", "caNRD_edit", "joint_reconstruction"))
  ggplot2::ggplot(d, ggplot2::aes(genotype, editing)) + ggplot2::geom_boxplot(outlier.size = 0.3) +
    ggplot2::facet_grid(source ~ celltype, scales = "free_y") + ggplot2::theme_bw() +
    ggplot2::labs(y = "editing (%)", x = "alternate allele count", title = paste("Monocyte-only site", site))
}
```

# Session info

```{r session}
sessionInfo()
```
